---
title: "NBA Salary Prediction"
author: "Eric Cheung"
format: html
latex: true
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| message: false
library(tidyverse)
library(scales)
```

```{r}
nba_sal <- read_csv("nba_2022-23_all_stats_with_salary.csv")
nba_dat <- nba_sal |> select(`Player Name`:Age, GP, MP, TRB:PTS)

nba_dat |>
  select(-`Player Name`) |>
  summarise(
    avg_salary = mean(Salary),
    avg_age = mean(Age),
    avg_GP = mean(GP),
    avg_MP = mean(MP),
    avg_TRB = mean(TRB),
    avg_AST = mean(AST),
    avg_STL = mean(STL),
    avg_BLK = mean(BLK),
    avg_TOV = mean(TOV),
    avg_PF = mean(PF),
    avg_PTS = mean(PTS)
  ) |>
  pivot_longer(
    cols = avg_salary:avg_PTS,
    names_to = "Averages",
    values_to = "Values"
  )

```

# Introduction

The average NBA player in the during the 2022-23 season made a salary of
8.4 million dollars while averaging 9.1 points, 3.5 rebounds, and 2.1
assists per game. I want to create some models that will attempt to
predict an NBA players salary. I will based these models solely on
statistics from the player's basic box score. The data I have is of 467
NBA players' salaries and their basic statistics including points per
game, assists per game, total rebound per game, etc.

------------------------------------------------------------------------

```{r}
nba_dat |> ggplot(aes(x = PTS, y = Salary)) + 
  geom_point(aes(color = Position)) + 
  geom_smooth(method = lm, se = FALSE) + 
  scale_y_continuous(labels = label_dollar()) +
  labs(x = "Points Per Game",
       title = "PPG vs Salary") + 
  theme_minimal()

cor(nba_dat$Salary, nba_dat$PTS)
```

There seems to be a general trend that players who average more points
per game make a higher salary. The correlation between those two
variables is quite high at 0.72 which indicates a strong linear
relationship. However, which position a player plays does not seem to
have an effect on the salary as there are highly paid player from all
five positions. Therefore I will not include the position variable in
any of my models so it won't introduce any unneeded variance. I believe
that points is one of the biggest on-court factor that contributes to a
players salary therefore I will include it in all of my models.

------------------------------------------------------------------------

```{r}
nba_dat |> ggplot(aes(x = MP, y = Salary)) + 
  geom_point(aes(color = Position)) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 3), se = FALSE) + 
  scale_y_continuous(labels = label_dollar()) +
  labs(x = "Minutes Played Per Game",
       title = "MP vs Salary") +
  theme_minimal()
```

There appears to be a polynomial relationship between minutes played and
salary as most highly paid players do not play past 35 minutes per game
with a few of the highest paid players are playing between 20-30 minutes
a night. I will model this relationship in my model with a polynomial
regression in one of the models

------------------------------------------------------------------------

```{r}
nba_dat |> 
  ggplot(aes(x = MP, y = PTS)) + 
  geom_point(aes(color = Position)) +
  geom_smooth(method = lm, formula = y ~ poly(x, 3), se = FALSE) + 
  scale_y_continuous(labels = label_dollar()) + 
  labs(title = "Points Per Game vs Minutes Played Per Game",
       x = "Minutes Played Per Game",
       y = "Points Per Game") + 
  theme_minimal()

```

There is a strong polynomial relationship between points per game and
minutes played per game. There has been a trend going on in the NBA
where star players are not playing over 35 minutes per game compared to
the early 2000s where stars would regularly play over 40 minutes. But
perhaps this trend will not play much of an effect on the salaries of
the players as they have not stopped growing ever since. I will model
this relationship through an interaction term between minutes played and
points per game.

------------------------------------------------------------------------

# The Models

### Linear

$$
\text{Salary} = \beta_0 + \beta_1 \cdot \text{PTS} + \beta_2 \cdot \text{TRB} + \beta_3 \cdot \text{AST} + \beta_4 \cdot \text{Age} \\
+ \beta_5 \cdot MP + \beta_6 \cdot MP^2 + \beta_7 \cdot MP^3 + \epsilon
$$

### Polynomial

$$
\text{Salary} = \beta_0 + \beta_1 \cdot \text{PTS} + \beta_2 \cdot \text{TRB} + \beta_3 \cdot \text{AST} + \beta_4 \cdot \text{Age} \\
+ \beta_5 \cdot \text{MP} + \beta_6 \cdot \text{MP}^2 + \beta_7 \cdot \text{MP}^3 + \epsilon
$$

### Interaction

$$
\text{Salary} = \beta_0 + \beta_1 \cdot \text{PTS} + \beta_2 \cdot \text{TRB} + \beta_3 \cdot \text{AST} + \\ \beta_4 \cdot \text{Age} + \beta_5 \cdot \text{MP} + \beta_6 \cdot \text{PTS} \times \text{MP} + \epsilon
$$

------------------------------------------------------------------------

```{r}
get_MSPE <- function(Y, Y_hat) {
  residuals <- Y - Y_hat
  resid_sq <- residuals^2
  SSPE <- sum(resid_sq)
  MSPE <- SSPE / length(Y)
  return(MSPE)
}
```

This is a function that will get the MSPE which is a measure of how well
a model performs for prediction.

```{r}
set.seed(2003125)
n <- nrow(nba_dat)
n_fold <- ceiling(n/10)
order_ids <- rep(1:10, times = n_fold)
order_ids <- order_ids[1:n]
shuffle <- sample.int(n)
shuffled_ids <- order_ids[shuffle]

dat_cv <- nba_dat
dat_cv$Salary <- log(dat_cv$Salary)
dat_cv$fold <- shuffled_ids

CV_MSPEs <- array(0, dim = c(10, 3))
colnames(CV_MSPEs) <- c("Linear", "Poly", "Interaction")

for (i in 1:10) {
  data_train <- filter(dat_cv, fold != i)
  data_valid <- filter(dat_cv, fold == i)
  
  lin_mod <- lm(Salary ~ PTS + TRB + AST + Age + MP, data = data_train)
  poly_mod <- lm(Salary ~ PTS + TRB + AST + Age + poly(MP,3), data = data_train)
  intera_mod <- lm(Salary ~ PTS + TRB + AST + Age + MP + PTS:MP, data = data_train)
  
  lin_pred <- predict(lin_mod, data_valid)
  poly_pred <- predict(poly_mod, data_valid)
  intera_pred <- predict(intera_mod, data_valid)
  
  Y_valid <- data_valid$Salary
  MSPE_lin <- get_MSPE(Y_valid, lin_pred)
  MSPE_poly <- get_MSPE(Y_valid, poly_pred)
  MSPE_intera <- get_MSPE(Y_valid, intera_pred)
  
  CV_MSPEs[i, 1] <- MSPE_lin
  CV_MSPEs[i, 2] <- MSPE_poly
  CV_MSPEs[i, 3] <- MSPE_intera
  
}
```

I used 10 fold cross validation to judge the performance of my three
models. This will produce 10 calculation of MSPE for each of the 3
models. I will use the average of the 10 MSPE to use as a measure of the
predictive power of each of the models.

------------------------------------------------------------------------

# Conclusion

```{r}
CV_dataframe <-  as.data.frame(CV_MSPEs) 
CV_dataframe |>
pivot_longer(cols = everything(),
names_to = "Model",
values_to = "CV_Error") |>
ggplot(aes(Model, CV_Error)) +
geom_boxplot() +
theme_bw() +
labs(y = "CV Error",
title = "CV Error with 10 Folds") +
  theme_minimal()

CV_dataframe |> colMeans()
```

The models I chose generally performed about the same as they have a
similar MSPE averaged out from the 10 fold cross validation. Seeing as
how they all perform about the same I will be using the linear model as
it is the most simple which will provide a good balance between model
bias and variance.

------------------------------------------------------------------------

```{r}
imaginary_player <- data.frame(PTS = 15.6, 
                               TRB = 7.3, 
                               AST = 3.6, 
                               Age = 26, 
                               MP = 30.2)

predicted_log_salary <- predict(lin_mod, newdata = imaginary_player)

predicted_salary <- exp(predicted_log_salary)
predicted_salary
```

So according to my model an imaginary player who is a solid starter
playing 30.2 minutes per night averaging 15.6 points, 7.3 rebounds, 3.6
assists, and aged 26 should be paid a salary of 11.4 million per year in
2022-23 money.
